---
title: "project2"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(readr)
library(ggplot2)
library(moments)
library(lubridate)
options(scipen = 9999)
options(warn = -1)
```

## importing data
```{r}
chi <- read_csv("chicago.csv")
ny <- read_csv("new-york-city.csv")
wash <- read_csv("washington.csv")
```
#We will check the summaries of each data set using sapply function applying summary statistics and glimpse

```{r}

sapply(list("Chicago" = chi,"New York" = ny,"Washington"  =  wash), glimpse)

sapply(list("Chicago" = chi,"New York" = ny,"Washington"  =  wash), summary)

```
## we can see that  ther's  Na  values so we need to handel it first
# let's check how many Na Values in each file for each variable  we gonna use the pipe '%>%'
```{r}
chi %>% summarise_all(funs(sum(is.na(.))))
ny %>% summarise_all(funs(sum(is.na(.))))
wash %>% summarise_all(funs(sum(is.na(.))))
```
# now let's prepare the date data to answer the questions

```{r}
#Created a function to extract_time hour, month and day from start_time of all three csv files

extract_time <- function(data){
  data$hour <- hour(data$`Start Time`)
  data$month <- month(data$`Start Time`)
  data$day <- weekdays(data$`Start Time`)
  return(data)
}
```
#Now let's have our three datasets chicago, newyork and washington on which we will perform the exploratory analysis and answer  the  questions
```{r}
chicago <- extract_time(chi)
newyork <- extract_time(ny)
washington <- extract_time(wash)

```
#now we can start  the Exploratory Data Analysis on the cleaned data sets "chicago, newyork and washington"

#Q1 Popular times of travel (i.e., occurs most often in the start time)
 first in chicago
```{r}

ggplot(chicago, aes(month)) + geom_histogram(bins = 6, color = I('white'))+  
  scale_x_continuous(breaks = seq(1,6,1))+ scale_y_continuous(breaks = seq(0,100000, 10000))+
  xlab("month") +  ylab("Count")+ ggtitle("Popular Months in Chicago")

```


## it's clearly that  Popular times of travel is in the summer exactly in JUN  with number of trips reaching upto 100000

seconed in newyork
```{r}

ggplot(washington, aes(month)) + geom_histogram(bins = 6, color = I('white'))+  
  scale_x_continuous(breaks = seq(1,6,1))+ scale_y_continuous(breaks = seq(0,100000, 10000))+
  xlab("month") +  ylab("Count")+ ggtitle("Popular Months in Washington")

```
## same as in chicago in JUN but  number of trips little below 70000 trips which means less than Chicago


third New York
```{r}
ggplot(newyork, aes(month)) + geom_histogram(bins = 6, color = I('white'))+  
  scale_x_continuous(breaks = seq(1,6,1))+ scale_y_continuous(breaks = seq(0,100000, 10000))+
  xlab("month") +  ylab("Count")+ ggtitle("Popular Months in New York City")

```
#The most popular month in NewYork City is month JUN  with number of trips reaching upto 75000 trips which is less than chicago

######
#What is the most common day of week?

in chicago
```{r}
ggplot(chicago, aes(day)) + geom_bar()+  
  xlab("Weekdays") +  ylab("Count")+ ggtitle("Popular Weekdays in Chicago")

#Monday and Tuesday are the most common days where the number of trips reach around 40000
```
in newyork

```{r}
ggplot(newyork, aes(day)) + geom_bar()+  
  xlab("Weekdays") +  ylab("Count")+ ggtitle("Popular Weekdays in New York City")

#Wednesday followed by Thursday are the most popular trips in Newy York City with number of trips reaching upto 50000.
```

in washington

```{r}
ggplot(washington, aes(day)) + geom_bar()+  
  xlab("Weekdays") +  ylab("Count")+ ggtitle("Popular Weekdays in Washington")

#Wednesday sees most trips in washington.
```
###
#What is the most common hour of day in all cities?

in chicago

```{r}
ggplot(chicago, aes(hour)) + geom_histogram(bins = 24, color = I('white'))+  
  scale_x_continuous(breaks = seq(0,23,1))+
  xlab("hour") +  ylab("count")+ ggtitle("Popular hours in Chicago")

#The morning peak hour in chicago id 8 am around 25000 and evening peak hour is 5 pm around 35000 trips.
```

in newyork

```{r}
ggplot(newyork, aes(hour)) + geom_histogram(bins = 24, color = I('white'))+  
  scale_x_continuous(breaks = seq(0,23,1))+
  xlab("hour") +  ylab("count")+ ggtitle("Popular hours in New York City")

#The morning peak hour in newyork is 8 am around 25000 and evening peak hour is 5-6 pm around 30000 trips.
```

in washington
```{r}
ggplot(washington, aes(hour)) + geom_histogram(bins = 24, color = I('white'))+  
  scale_x_continuous(breaks = seq(0,23,1))+
  xlab("hour") +  ylab("count")+ ggtitle("Popular hours in Washington")


#The trend in washington is slightly different, where the peak hours start from 1 am and reaches maximum at 8 am with 30000 trips.
#Although evening peak hour is around 5 pm but the number of trips is around 10000 which is much lesser than other cities.
#Note: Limits were set in x-axis for better view of the ranges in x-axis


```
##Q2 Trip duration

```{r}


#What is the gender wise median travel time for users in different cities ?

summary(chicago$`Trip Duration`)

#Median trip duration in chicago is 670.0 seconds and Mean trip duration is 936.2 seconds

ggplot(data = subset(chicago, !is.na(Gender)), aes("Gender",`Trip Duration`, fill = Gender)) + geom_boxplot()+  
  scale_y_continuous(limits = c(0,2000))+
  xlab("Gender") +  ylab("Time in Seconds ")+ ggtitle("Trip duration in Chicago")


summary(newyork$`Trip Duration`)

#Median trip duration in newyork is 609.0 seconds and Mean trip duration is 899.7 seconds

ggplot(data = subset(newyork, !is.na(Gender)), aes("Gender",`Trip Duration`, fill = Gender)) + geom_boxplot()+  
  scale_y_continuous(limits = c(0,2000))+
  xlab("Gender") +  ylab("Time in Seconds ")+ ggtitle("Trip duration in newyork")

#The median trip duration of females is more than their male counterparts.
#Note: Limits were set in y-axis for better view of the ranges in y-axis
#Gender Legend is used to view the median trip duration of both gender types in boxplots. 

#Q3 User info

#What are the counts of each gender (only available for NYC and Chicago)?

chicago %>% group_by(Gender) %>% filter(!is.na(Gender))%>%
  summarise(total = length(Gender))%>%
  ggplot(aes(Gender, total)) + geom_bar(stat = 'identity')+
  xlab("Gender") +  ylab("Count ")+ ggtitle("Counts of each gender in chicago")


newyork %>% group_by(Gender) %>% filter(!is.na(Gender))%>%
  summarise(total = length(Gender))%>%
  ggplot(aes(Gender, total)) + geom_bar(stat = 'identity')+
  xlab("Gender") +  ylab("Count ")+ ggtitle("Counts of each gender in new york")


#Females are lesser in number as compared to Males in chicago and newyork
#Note: library dplyr was used to summarise the data
#observations where gender were NA were filtered out


#What are the earliest, most recent, most common year of birth (only available for NYC and Chicago)?

chicago %>% group_by(Gender,`Birth Year`) %>% filter(!is.na(Gender)) %>% 
  summarise(total = length(`Birth Year`))%>%
  ggplot(aes(`Birth Year`, total)) + geom_bar(stat = 'identity', fill = 'orange')+
  coord_trans(y = 'sqrt')+
  facet_grid(Gender~.)+
  scale_x_continuous(limits = c(1930,2017), breaks = seq(1930,2007,1))+
  theme(axis.text.x = element_text(angle = 60))+
  xlab("Birth Year") +  ylab("Count")+ ggtitle("Trip duration in chicago")


newyork %>% group_by(Gender,`Birth Year`) %>% filter(!is.na(Gender)) %>% 
  summarise(total = length(`Birth Year`))%>%
  ggplot(aes(`Birth Year`, total)) + geom_bar(stat = 'identity', fill = 'orange')+
  coord_trans(y = 'sqrt')+
  facet_grid(Gender~.)+
  scale_x_continuous(limits = c(1930,2017), breaks = seq(1930,2007,1))+
  theme(axis.text.x = element_text(angle = 60))+
  xlab("Birth Year") +  ylab("Count")+ ggtitle("Trip duration in newyork")


#The user count increases with increase in Birth Year, with highest number of users belong to 1988-89
#Note: Limits were set in x-axis for removing the outlier years and male and female facet grids were used.

```
```

